Legends for supplementary material

### 2019-04-24 Manuscript/Figures/figure_legends.Rmd
editted legends
added figure/table legends

**Figure 1: Model Construction.** An illustration of the proposed model construction given cell line-specific Hi-C matrices. The contact matrix above is a visualization of the interconnectivity of chromosome 14 at 100 kb resolution, for the GM12878 cell line, on the intervals from 20 mb to 100 mb. The boundary coordinates of topologically association domains (TADs) were identified using the Arrowhead algorithm. To account for the uncertainty in TAD boundary detection, each boundary point was flanked on both sides by resolution-specific genomic regions (bins). The response vector $Y$, used for classification, was determined by whether a genomic bin overlapped with a flanked boundary region (Y=1) or not (Y=0). The spatial relationship between positional coordinates of various functional genomic annotations, obtained from ENCODE, and the genomic bins were then used to define the feature space of the models.

**Figure 2: Ensemble Framework/Model Building Pipeline.** A diagram of the model building pipeline given a combination of inputs from the ensemble framework. A particular ensemble system included the total number of genomic bins, labeled as having been overlapped with a flanked TAD boundary or not, as well as a feature space composed of 1 of 3 different predictor-type specific annotations. The data was then split into a 7:3 training:testing set ratio. A re-sampling technique was applied to the training set, followed by an elastic-net regularization to reduce the feature space. Lastly, a random forest classification model was performed on the re-sampled and reduced training set and validated on the held out testing set common to all ensemble systems in the framework.

**Figure 3. The three predictor types considered when quantifying the association between TAD boundaries and genomic annotations.** The _overlap count_ (OC) predictors consider the total number of genomic features that overlapped with each genomic bin. The _overlap percent_ (OP) predictors were calculated by dividing the sum of all feature widths within each specific bin by the total bin width. The _distance_ ($log_{2}$ transformed) predictors were calculated by measuring the distance (in bases) from the center of each genomic bin to the center of the nearest genomic feature.

**Figure 4. Data summaries in the form of barplots for the GM12878 (red) and K562 (blue) cell lines.** (A) The total number of genomic bins created by partitioning the linear genome into unit intervals corresponding to the resolution of the relevant domain data. (B) Subsequent imbalance ratios, defined as the ratio of the number of bins that overlapped with a flanked TAD boundary to the number of bins that did not overlap with a flanked TAD boundary.

**Figure 7. Evaluating effect of class imbalance on model performance.**

>>>>

**Figure 1. Model Construction.** An illustration of the proposed model construction given cell line-specific Hi-C matrices. The contact matrix above is a visualization of the interconnectivity of chromosome 14 at 100 kb resolution, for the GM12878 cell line, on the intervals from 20 mb to 100 mb. The boundary coordinates of topologically association domains (TADs) were identified using the Arrowhead algorithm. To account for the uncertainty in TAD boundary detection, each boundary point was flanked on both sides by resolution-specific genomic regions (bins). The response vector $Y$, used for classification, was determined by whether a genomic bin overlapped with a flanked boundary region (Y=1) or not (Y=0). The spatial relationship between positional coordinates of various functional genomic annotations, obtained from ENCODE, and the genomic bins were then used to define the feature space of the models.

**Figure 2. Ensemble Framework.** A diagram of the ensemble framework detailing the various combination of inputs used in the model building pipeline. A given system within the ensemble framework included the total number of genomic bins, labeled as having been overlapped with a flanked TAD boundary or not, as well as a feature space composed of 1 of 3 different predictor-type specific annotations. The data was then split into a 7:3 training:testing set ratio. A re-sampling technique was applied to the training set, followed by an elastic-net regularization to reduce the feature space. Lastly, a random forest classification model was performed on the re-sampled and reduced training set and validated on the held out testing set common to all ensemble systems in the framework.

**Figure 3. The three predictor types.** Three separate types of predictors were considered when quantifying the association between TAD boundaries and genomic annotations. Each predictor type was used to build the feature space of specific models in the ensemble framework. The _overlap count_ (OC) predictors consider the total number of genomic features that overlapped with each genomic bin. The _overlap percent_ (OP) predictors were calculated by dividing the sum of all feature widths within each specific bin by the total bin width. The _distance_ ($log_{2}$ transformed) predictors were calculated by measuring the distance (in bases) from the center of each genomic bin to the center of the nearest genomic feature. [???should I include ABC here?]

**Figure 4. Resolution-specific data summaries.** (A) The total number of genomic bins created by partitioning the linear genome into unit intervals corresponding to the resolution of the relevant domain data for the GM12878 (red) and K562 (blue) cell line. (B) Subsequent imbalance ratios created when labelling bin class, defined as the ratio of the number of bins that overlapped with a flanked TAD boundary to the number of bins that did not overlap with a flanked TAD boundary.

**Figure 5. Comparing ensemble system specific model performances for GM12878.** Performances, separated by F1-Score and MCC, of random forest classification models built on each of the 3 predictor types: OC (A), OP (B), and Distance (C). Within each predictor type, and across resolution, F1-Scores and MCCs metrics were recorded for models subsequently implemented on re-sampled data using the 4 re-sampling techniques: None (red), ROS (green), RUS (blue), and SMOTE (purple).

**Figure 8. Predictive importance of genomic annotations for random forest models implemented on SMOTE resampled data with distance-type predictors for GM12878.** Importance was rescaled to values between 0 and 100, and discretized. The importance was subsequently sorted according to the most predictive annotations associated with the 10 kb resolution. Annotations with higher values of predictive importance are denoted in dark red, while annotations that were regularized in the pipeline for specific resolutions are denoted in grey.

**Figure 6. Comparing category-specific random forest models implemented on SMOTE re-sampled data with distance-type predictors for GM12878.** Performances, separated by F1-Score and MCC, were compared for models using specific categories of annotations as the feature space including: histone modifications (red), chromatin state segmentations (green), transcription factor binding sites, and a feature space containing all 3 types combined (purple).

**Figure 7. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 10 kb resolution, on GM12878.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.

### 2019-04-24 Manuscript/Supplementary/supplementary_legends.Rmd
added figure/table legends

**Supplementary Table 4. Hi-C Data.** A complete list of the Hi-C data libraries used to obtain the locations of topologically associating domains.

**Supplementary Table 1. Genomic Annotations.** A complete list of the genomic annotations used to build the feature space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic features, "Available Cell Type" - the name of the cell type in which the annotation was available, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

**Supplementary Figure 1. Distance Densities across cell lines and resolutions.** Density curves of distances (in millions of base pairs) from genomic bins to flanked TAD boundaries for 10 kb (A), 25 kb (B), 50 kb (C), and 100 kb (D) data resolutions across the GM12878 (left column) and K562 (right column) cell lines. The untransformed distance densities are filled in red, while the $log_{2}$ tranformed distance densities are filled in blue.

>>>>

**Supplementary Table 4. Hi-C data.** A complete list of the Hi-C data libraries used to obtain the locations of topologically associating domains.

**Supplementary Figure 14. Number of TADs called throughout the genome.** TADs were called throughout each of the 22 autosomal chromosomes for cell lines GM12878 (red) and K562 (blue) using the Arrowhead algorithm [@durand2016juicebox], across each data resolution including 10 kb, 25 kb, 50 kb, and 100 kb (A-D) respectively.

**Supplementary Table 8. Summary statistics for TAD lengths.** The following summary statistics were recorded regarding lengths of TADs that were called and delineated into TAD boundaries including the minimum, 1st quartile, median, mean, 3rd quartile, and maximum TAD lengths (in basepairs). The summaries are provided for each cell line, resolution, and within each chromosome. [???should I leave space for 9th chromosome on k562 where no TADs were called?]

**Supplementary Table 2. Genomic annotations.** A complete list of the genomic annotations used to build the feature space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic features, "Available Cell Type" - the name of the cell type in which the annotation was available, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

**Supplementary Figure 1. Distance densities from centers of genomic bins to closest annotation.** Density curves of the distances (in millions of base pairs) from the center of genomic bins to the center of the nearest respective annotation for the GM12878 (left column) and K562 (right column) cell lines respectively. Each curve represents the raw (red) and $log_{2}$ transformed (blue) distance profiles for each annotation, across each of the 4 resolutions (A-D).

**Supplementary Figure 2. Clustered correlogram for genomic annotations at the 10 kb resolution for the GM12878 cell line.** Pairwise correlations between the total number of instances where regions defined by genomic annotations overlapped with flanked TAD boundaries on the linear genome. Positive correlations are denoted in red, while negative correlations are denoted in blue. The upper triangular matrix of correlations was clustered using Ward's method. Delineated regions of high correlations indicate genomic annotations that are colocalized at TAD boundaries.

**Supplementary Figure 3. Clustered correlogram for genomic annotations at the 10 kb resolution for the K562 cell line.** Pairwise correlations between the total number of instances where regions defined by genomic annotations overlapped with flanked TAD boundaries on the linear genome. Positive correlations are denoted in red, while negative correlations are denoted in blue. The upper triangular matrix of correlations was clustered using Ward's method. Delineated regions of high correlations indicate genomic annotations that are colocalized at TAD boundaries.

**Supplementary Figure 4. Clustered correlogram for genomic annotations at the 25 kb resolution for GM12878 cell line.** Pairwise correlations between the total number of instances where regions defined by genomic annotations overlapped with flanked TAD boundaries on the linear genome. Positive correlations are denoted in red, while negative correlations are denoted in blue. The upper triangular matrix of correlations was clustered using Ward's method. Delineated regions of high correlations indicate genomic annotations that are colocalized at TAD boundaries.

**Supplementary Figure 5. Clustered correlogram for genomic annotations at the 25 kb resolution for the K562 cell line.** Pairwise correlations between the total number of instances where regions defined by genomic annotations overlapped with flanked TAD boundaries on the linear genome. Positive correlations are denoted in red, while negative correlations are denoted in blue. The upper triangular matrix of correlations was clustered using Ward's method. Delineated regions of high correlations indicate genomic annotations that are colocalized at TAD boundaries.

**Supplementary Figure 6. Clustered correlogram for genomic annotations at the 50 kb resolution for GM12878 cell line.** Pairwise correlations between the total number of instances where regions defined by genomic annotations overlapped with flanked TAD boundaries on the linear genome. Positive correlations are denoted in red, while negative correlations are denoted in blue. The upper triangular matrix of correlations was clustered using Ward's method. Delineated regions of high correlations indicate genomic annotations that are colocalized at TAD boundaries.

**Supplementary Figure 7. Clustered correlogram for genomic annotations at the 50 kb resolution for the K562 cell line.** Pairwise correlations between the total number of instances where regions defined by genomic annotations overlapped with flanked TAD boundaries on the linear genome. Positive correlations are denoted in red, while negative correlations are denoted in blue. The upper triangular matrix of correlations was clustered using Ward's method. Delineated regions of high correlations indicate genomic annotations that are colocalized at TAD boundaries.

**Supplementary Figure 8. Clustered correlogram for genomic annotations at the 100 kb resolution for GM12878 cell line.** Pairwise correlations between the total number of instances where regions defined by genomic annotations overlapped with flanked TAD boundaries on the linear genome. Positive correlations are denoted in red, while negative correlations are denoted in blue. The upper triangular matrix of correlations was clustered using Ward's method. Delineated regions of high correlations indicate genomic annotations that are colocalized at TAD boundaries.

**Supplementary Figure 9. Clustered correlogram for genomic annotations at the 100 kb resolution for the K562 cell line.** Pairwise correlations between the total number of instances where regions defined by genomic annotations overlapped with flanked TAD boundaries on the linear genome. Positive correlations are denoted in red, while negative correlations are denoted in blue. The upper triangular matrix of correlations was clustered using Ward's method. Delineated regions of high correlations indicate genomic annotations that are colocalized at TAD boundaries.

**Supplementary Table 7. Optimal alpha and lambda penalization parameters.** The tuned alpha and lambda penalization parameter terms after performing 10-fold cross validated elastic-net regularization specific to cell line, resolution, re-sampling technique, and predictor type.

**Supplementary Figure 10. Comparing ensemble system specific model performances for K562.** Performances of random forest classification models built on each of the 3 predictor types: OC (A), OP (B), and Distance (C). Within each predictor type, and across resolution, F1-Scores and MCCs metrics were recorded for models subsequently implemented on re-sampled data using the 4 re-sampling techniques: None (red), ROS (green), RUS (blue), and SMOTE (purple).

**Supplementary Table 4. Ranking predictive importance between predictor types for annotation-specific models for GM12878.** Annotation-specific random forest classification models were implemented on SMOTE resampled data, across resolution. Each model contained each of the 3 predictor types for one genomic annotation. Predictive importance was rescaled to values between 0 and 100. The predictor type with importance value of 100 indicates it was the most informative predictor type out of the 3.

**Supplementary Table 5. Ranking predictive importance between predictor types for annotation-specific models for K562.** Annotation-specific random forest classification models were implemented on SMOTE resampled data, across resolution. Each model contained each of the 3 predictor types for one genomic annotation. Predictive importance was rescaled to values between 0 and 100. The predictor type with importance value of 100 indicates it was the most informative predictor type out of the 3.

**Supplementary Figure 15. Predictive importance of genomic annotations for random forest models implemented on SMOTE resampled data with distance-type predictors for K562.** Importance was rescaled to values between 0 and 100, and discretized. The importance was subsequently sorted according to the most predictive annotations associated with the 10 kb resolution. Annotations with higher values of predictive importance are denoted in dark red, while annotations that were regularized in the pipeline for specific resolutions are denoted in grey.

**Supplementary Figure 11. Comparing category-specific random forest models implemented on SMOTE re-sampled data with distance-type predictors for K562.** Performances, separated by F1-Score and MCC, were compared for models using specific categories of annotations as the feature space including: histone modifications (red), chromatin state segmentations (green), transcription factor binding sites, and a feature space containing all 3 types combined (purple).

**Supplementary Figure 12. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 25 kb resolution, on GM12878.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.

**Supplementary Figure 13. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 50 kb resolution, on GM12878.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.

**Supplementary Figure 14. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 100 kb resolution, on GM12878.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.

**Supplementary Figure 15. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 10 kb resolution, on K562.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.

**Supplementary Figure 16. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 25 kb resolution, on K562.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.

**Supplementary Figure 17. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 50 kb resolution, on K562.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.

**Supplementary Figure 18. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 100 kb resolution, on K562.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.
